---
title: "Time Series Analysis Lab B"
author: "Thijs Quast (thiqu264)"
date: "9/23/2019"
output: pdf_document
toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage
# Assignment 1. Computations with simulated data
## a
Generate 1000 observations from AR(3) process with phi1 = 0.8, phi2 = -0.2, phi3 = 0.1. Use these data and the definition of PACF to compute phi33 from the sample, i.e. write your own code that performs linear regressions on necessarily lagged variables and then computes an appropriate correlation. Compare the result with the output of function pacf() and with the theoretical value of
phi33.

$$ X_t = 0.8X_{t-1} - 0.2X_{t-2} + 0.1X_{t-3} + w_t$$ 
$$ W_t distributed N(0,1)$$

```{r}
# pacf using linear regression
set.seed(12345)
data=arima.sim(list(ar=c(0.8,-0.2,0.1)), n=1000)
data1=ts.intersect(x=data, x1=lag(data,-1), x2=lag(data,-2), x3=lag(data,-3), dframe = T)

res1=lm(x~x1+x2,data=data1)
res2=lm(x3~x2+x1,data=data1)
r1=residuals(res1)
r2=residuals(res2)
cor(cbind(r1,r2))
```

```{r}
# pacf function
g=pacf(data)
g$acf[3]
```


```{r}
# Theoretical pacf
theoretical_pacf <- ARMAacf(ar=c(0.8, -0.2, 0.1), lag.max = 30, pacf = TRUE)
theoretical_pacf[3]
```

```{r}
# Comparing in plots
pacf(data, main="Sample PACF")
plot(theoretical_pacf, xlab = "Lag", ylab = "Theoretical PACF", type = "h",
     main = "Theoretical PACF")
abline(h=0)
```

## b
```{r}
set.seed(12345)
AR2 <- arima.sim(model = list(ar=c(0.8, 0.1)), n = 100)

# Yule-Walker
AR2_yw <- ar(x = AR2 ,method = "yw")

# Conditional Least Squares
AR2_cls <- arima(x = AR2, order = c(2,0,0), method = "CSS")

# ML
AR2_ML <- arima(x = AR2, order = c(2,0,0), method = "ML")
```

```{r}
AR2_yw
```

```{r}
AR2_cls
```

```{r}
AR2_ML
```


```{r}
# Theoretical value phi2

```

```{r}
confint(AR2_ML)
```


## c
```{r}
omega <- c(0.3, rep(0, 11), 0.6)
sAR <- arima.sim(model = list(order=c(0,0,13), ma=omega), n = 200)
```

```{r}
# Sample ACF and Sample PACF
acf(sAR, main="Sample ACF")
pacf(sAR, main="Sample PACF")
```

## d
```{r}
library(astsa)
x <- arima.sim(n = 200, list(order(0,0,13), ma=c(0.3, rep(0, 10), 0.6, 0.18)))
fit <- sarima(xdata = AR2, 0, 0, 1, 0, 0, 1, 12)
fit
```

```{r}
sarima.for(xdata = sAR, 12, 0, 0, 1, 0, 0, 1, 12)
```

```{r}
# Forecast, using kernlab package
library(kernlab)
x <- c(1:200)
x_test <- c(0:230)
fit_gausspr <- gausspr(x = x, y=sAR)

# Add noise to predictions
#noise <- rnorm(30)
y_pred <- predict(fit_gausspr, x_test)

```

```{r}
ts.plot(sAR)
lines(y_pred, col="red")
```

## e
```{r}
set.seed(12345)
ARMA_11 <- arima.sim(model = list(order=c(1,0,1), ar=0.7, ma=0.5), n = 50)
ARMA_11_fit <- arima(x = ARMA_11[1:40], order = c(1,0,1))
```

```{r}
prediction <- predict(ARMA_11_fit, n.ahead = 10)
ts.plot(ts(ARMA_11[1:50]), prediction$pred, col=c(1:2), type="o")


U = prediction$pred+2*prediction$se 
L = prediction$pred-2*prediction$se
xx = c(time(U), rev(time(U)));  yy = c(L, rev(U))
polygon(xx, yy, border = 8, col = gray(.6, alpha = .2))
lines(prediction$pred, type="p", col=2)
```

# Assignment 2
## a
```{r}
# Chicken
x_t <- chicken
ts.plot(x_t)
title(main = "Chicken")
```

```{r}
# Chicken
par(mfrow=c(2,2), oma=c(0,0,2,0))
acf(x_t, lag.max = 40)
pacf(x_t, lag.max = 40)
acf(diff(x_t, lag = 1), lag.max = 40)
pacf(diff(x_t, lag = 1), lag.max = 40)
title(main = "ACF and PACF for Chicken", outer = TRUE)
```

Based on the fact that there is high ACF, but PACF seems to be low, in addition there seems to be seasonality for 12 lags. So, I'd say we go for an ARIMA(1,1,0)x(1,0,0)12

## b
```{r}
#so2
x_t <- so2
ts.plot(x_t)
title(main = "so2")
```

```{r}
# so2
par(mfrow=c(2,2), oma=c(0,0,2,0))
acf(x_t)
pacf(x_t)
acf(diff(x_t, lag = 1))
pacf(diff(x_t, lag = 1))
title(main = "ACF and PACF for so2", outer = TRUE)
```


```{r}
# EQCount
x_t <- EQcount
ts.plot(x_t)
title(main = "EQcount")
```


```{r}
# EQCount
par(mfrow=c(2,2), oma=c(0,0,2,0))
acf(x_t)
pacf(x_t)
acf(diff(x_t, lag = 1),lag.max=40)
pacf(diff(x_t, lag = 1),lag.max=40)
title(main = "ACF and PACF for EQcount", outer = TRUE)
```

```{r}
x_t <- HCT
ts.plot(x_t)
title(main = "HCT")
```

```{r}
# EQCount
par(mfrow=c(2,2), oma=c(0,0,2,0))
acf(x_t)
pacf(x_t)
acf(diff(x_t, lag = 1))
pacf(diff(x_t, lag = 1))
title(main = "ACF and PACF for HCT", outer = TRUE)
```

# Assignment 3
## a
```{r}
ts.plot(oil)
title(main = "Oil price time series")
```

```{r}
# Non-stationary, so use differencing:
ts.plot(diff(oil))
```

```{r}
# Variance is not constant everywhere, so:
ts.plot(diff(log(oil)))
```
Variance is much more equal now across the entire time series. Look at the y-scale of plot above.

```{r}
doil <- diff(log(oil))
```

```{r}
library(tseries)
```

```{r}
adf.test(doil)
```

```{r}
acf2(doil)
```

```{r}
library(TSA)
eacf(doil)
```

```{r}
# ARMA(1,1)
fit1 <- sarima(doil, p = 1, d = 1, q = 1)
hist(residuals(fit1$fit), breaks = 25)
runs(residuals(fit1$fit))
```

```{r}
# ARMA(0,3)
fit2 <- sarima(doil, p = 0, d = 1, q = 3)
hist(residuals(fit2$fit), breaks = 25)
runs(residuals(fit2$fit))
```

```{r}
logoil <- arima(log(oil), order = c(3,0,3))
prediction <- predict(logoil, n.ahead=20)
```


```{r}
ts.plot(oil, exp(prediction$pred), col=c(1:2))

U <- exp(prediction$pred + 2*prediction$se)
L <- exp(prediction$pred - 2*prediction$se)

xx = c(time(U), rev(time(U)));  yy = c(L, rev(U))
polygon(xx, yy, border = 8, col = gray(.6, alpha = .2))
lines(prediction$pred, type="p", col=2)
```

## b
```{r}
ts.plot(unemp)
title(main = "Unemployment Rates")
```

```{r}
# Non-stationary, so use differencing:
ts.plot(diff(unemp))
```

```{r}
# Variance is not constant everywhere, so:
ts.plot(diff(log(unemp)))
```

```{r}
dunemp <- diff(log(unemp))
```

```{r}
library(tseries)
```

```{r}
adf.test(dunemp)
```

```{r}
acf2(dunemp)
```

```{r}
library(TSA)
eacf(dunemp)
```


```{r}
# ARMA(1,1)
fit1 <- sarima(dunemp, p = 1, d = 1, q = 1)
hist(residuals(fit1$fit), breaks = 25)
runs(residuals(fit1$fit))
```

```{r}
# ARMA(2,2)
fit2 <- sarima(dunemp, p = 2, d = 1, q = 2)
hist(residuals(fit2$fit), breaks = 25)
runs(residuals(fit2$fit))
```

```{r}
# ARMA (3,5)
fit3 <- sarima(dunemp, p = 3, d = 1, q = 5)
hist(residuals(fit3$fit), breaks = 25)
runs(residuals(fit3$fit))
```



```{r}
# ARMA (3,5)
logunemp <- arima(log(unemp), order = c(3,0,5))
prediction <- predict(logunemp, n.ahead=20)
```


```{r}
ts.plot(unemp, exp(prediction$pred), col=c(1:2))

U <- exp(prediction$pred + 2*prediction$se)
L <- exp(prediction$pred - 2*prediction$se)

xx = c(time(U), rev(time(U)));  yy = c(L, rev(U))
polygon(xx, yy, border = 8, col = gray(.6, alpha = .2))
lines(prediction$pred, type="p", col=2)
```