---
title: "Lab3"
author: "Thijs Quast (thiqu264)"
date: "10/7/2019"
output: pdf_document
toc: TRUE
---
\newpage
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1
```{r}
set.seed(12345)
```

```{r}
transition <- function(z_t){
    action <- sample((0:2), 1)
    z_t <- rnorm(n = 1, mean = (z_t + action), sd = 1)
    
    return(z_t)
  }

```

```{r}
emission <- function(z_t){
  action <- sample((-1:1), 1)
  x_t <- rnorm(n = 1, mean = (z_t + action), sd = 1)
  
  return(x_t)
}
```

```{r}
weights <- function(observation, mean, sd){
  weight <- (dnorm(observation, mean = mean, sd = sd) + 
    dnorm(observation, mean = mean-1, sd = sd) +
    dnorm(observation, mean = mean+1, sd = sd))/3
  
  return(weight)
}
```


```{r}
z_t <- c()
x_t <- c()
z_t[1] <- runif(1, 0, 100)

for (i in 2:100){
  z_t[i] <- transition(z_t = z_t[i-1])
}

for (i in 1:100){
  x_t[i] <- emission(z_t = z_t[i])
}
```


```{r}
particle_filter <- function(observations, M, Time, sd, correction=TRUE){
  
  particles <- matrix(0, nrow = M, ncol = Time)
  
  # Initialization, using initial model:
  initialization <- runif(n = M, min = 0, max = 100)
  W <- matrix(data = 0, nrow = M, ncol = Time)
  
  # Prediction:
  for (t in 1:Time){ # For every timestep
    if (t==1){
      for (m in 1:M){ # Number of particles
      particles[m,t] <- transition(z_t = initialization[m])
      W[m,t] <- weights(observation = observations[m], mean = particles[m,t], sd = sd)
      
      W[,t] <- W[,t]/sum(W[,t])
      
      if (correction == TRUE){
        particles[,t] <- sample(particles[,t], M, replace=TRUE, prob=W[,t])
      } else {
        particles[,t] <- particles[,t]
      }
      
      }
    } else {
      for (m in 1:M){ # Number of particles
        particles[m,t] <- transition(z_t = particles[m, (t-1)])
        W[m,t] <- weights(observation = observations[m], mean = particles[m,t], sd = sd)
      
        W[,t] <- W[,t]/sum(W[,t])
      
        if (correction == TRUE){
          particles[,t] <- sample(particles[,t], M, replace=TRUE, prob=W[,t])
        } else {
          particles[,t] <- particles[,t]
      }
      }
    }
  }
  return(particles)
}
```

```{r}
# Creating the particle filter
filter <- particle_filter(observations = x_t, M = 100, Time = 100, sd = 1)
```

```{r}
library(ggplot2)
df <- as.data.frame(filter)
colnames(df) <- as.character(c(1:100))
```

```{r}
# Function to plot particle filter at different timesteps
plot_particle_filter <- function(data, timestep){
  plot <- ggplot(data = data, aes(x = data[,timestep])) + geom_density(colour="black")
  plot <- plot + xlab(paste("Time", as.character(timestep)))
  plot <- plot + geom_vline(aes(xintercept = data[,timestep], col="Corrected Particles"), alpha=0.20)
  plot <- plot + geom_vline(aes(xintercept = mean(data[,timestep]), col="Expected location"))
  plot <- plot + geom_vline(aes(xintercept = z_t[timestep], col="True location"))
  plot <- plot + ggtitle(paste("Particle filter at time", timestep))
  return(plot)
}
```

```{r}
plot_particle_filter(data = df, timestep = 2)
plot_particle_filter(data = df, timestep = 15)
plot_particle_filter(data = df, timestep = 85)
plot_particle_filter(data = df, timestep = 100)
```


# Question 2

```{r}
# Emmision model with sd = 5
emission_5 <- function(z_t){
  action <- sample((-1:1), 1)
  x_t <- rnorm(n = 1, mean = (z_t + action), sd = 5)
  
  return(x_t)
}
```

```{r}
x_t_5 <- c()

for (i in 1:100){
  x_t_5[i] <- emission_5(z_t = z_t[i])
}

```

```{r}
filter_5 <- particle_filter(observations = x_t_5, M = 100, Time = 100, sd = 5)
```

```{r}
library(ggplot2)
df_5 <- as.data.frame(filter_5$corrected_particles)
colnames(df_5) <- as.character(c(1:100))
```

```{r}
plot_particle_filter(data = df_5, timestep = 1)
plot_particle_filter(data = df_5, timestep = 15)
plot_particle_filter(data = df_5, timestep = 85)
plot_particle_filter(data = df_5, timestep = 100)
```


```{r}
# Emmision model with sd = 50
emission_50 <- function(z_t){
  action <- sample((-1:1), 1)
  x_t <- rnorm(n = 1, mean = (z_t + action), sd = 50)
  
  return(x_t)
}
```

```{r}
x_t_50 <- c()

for (i in 1:100){
  x_t_50[i] <- emission_50(z_t = z_t[i])
}

```

```{r}
filter_50 <- particle_filter(observations = x_t_50, M = 100, Time = 100, sd = 50)
```

```{r}
library(ggplot2)
df_50 <- as.data.frame(filter_50$corrected_particles)
colnames(df_50) <- as.character(c(1:100))
```

```{r}
plot_particle_filter(data = df_50, timestep = 1)
plot_particle_filter(data = df_50, timestep = 15)
plot_particle_filter(data = df_50, timestep = 85)
plot_particle_filter(data = df_50, timestep = 100)
```
Generally, as one increases the the standard deviation of the emission model the histogram of the particles is distributed over a wider deviation. 

# Question 3

```{r}
# Function to plot particle filter at different timesteps
plot_particle_filter2 <- function(data, timestep){
  plot <- ggplot(data = data, aes(x = data[,timestep])) + geom_histogram(binwidth = 1, colour="black")
  plot <- plot + xlab(paste("Time", as.character(timestep)))
  plot <- plot + geom_vline(aes(xintercept = data[,timestep], col="Particles"), alpha=0.20)
  plot <- plot + geom_vline(aes(xintercept = mean(data[,timestep]), col="Expected location"))
  plot <- plot + geom_vline(aes(xintercept = z_t[timestep], col="True location"))
  plot <- plot + ggtitle(paste("Particle filter at time", timestep))
  return(plot)
}
```


```{r}
df <- as.data.frame(filter$particles)
colnames(df) <- as.character(c(1:100))
```

```{r}
plot_particle_filter2(data = df, timestep = 1)
plot_particle_filter2(data = df, timestep = 15)
plot_particle_filter2(data = df, timestep = 85)
plot_particle_filter2(data = df, timestep = 100)
```

The difference with the correction in the particles filter is that without a correction the plot shows way more particles, distributed over a wider range. This is because in the correction the filter will select the ones with higher weights multiple times.